---
title: "Lab 08 Replication"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This is an replication notebook for Lab-08 from the course SOC 4650/5650: Introduction to GISc.

## Load Dependencies
The following code loads the package dependencies for our analysis:

```{r package-load}
# tidyverse packages
library(ggplot2)       # data plotting
library(dplyr)         # data wrangling
library(readr)         # work with csv files

# other packages
library(ggthemes)      # map theme
library(here)          # file path management
library(RColorBrewer)  # color ramps
library(sf)            # spatial data tools
```

## Load Data
We'll use one data set from the `stlData` package and one from this week's mini data release to practice table joins: census tract geometric data and the estimated percent of residents per tract with diabetes. Both are added here:

```{r load-data}
library(stlData)
stlTracts <- stl_sf_tracts
diabetes <- read_csv(here("data", "STL_HEALTH_Diabetes.csv"))
```

## Part 2
### Question 2
Before completing the join, we need to both identify the proper identification variables to base the join on and make sure they are the same type. In `stlTracts`, like all census geography data, the `GEOID` variable is the standard variable for completing joins. We test it using the `class()` function:

```{r assess-tracts-geoid}
class(stlTracts$GEOID)
```

In `diabetes`, the `tract_fips` variable looks similar to `GEOID`. We test it using the same `class()` function:

```{r assess-diabetes-geoid}
class(diabetes$tract_fips)
```

Since we have one character and one numeric variable, we need to convert one of them to be the other's type. Here, I'll convert the `tract_fips` variable to character using `mutate()` combined with `as.character()`:

```{r convert-diabetes-geoid}
diabetes <- mutate(diabetes, tract_fips = as.character(tract_fips))
```

Now we have identically formatted variables to base the join off of.

### Question 3
Next, we use `left_join()` to complete the table join:

```{r table-join}
stlDiabetes <- left_join(stlTracts, diabetes, by = c("GEOID" = "tract_fips"))
```

We now have a single table named `stlDiabetes` that has both the geometric data and the tabular data combined. With these data in a single table, we can proceed to mapping diabetes rates.

### Question 4
The following code produces a `ggplot` map of the diabetes rates data. We use `RColorBrewer` for a palatte and `ggthemes` for its clean `theme_map()`. Finally, we use the `labs()` function to apply appropriate labeling.

```{r map-diabetes}
ggplot() +
  geom_sf(data = stlDiabetes, mapping = aes(fill = estimate), color = NA) +
  scale_fill_distiller(palette = "RdPu", trans = "reverse", name = "Percent") +
  theme_map(base_size = 14) +
  theme(legend.position = "right") +
  labs(
    title = "Estimated Prevalence of \nDiabetes in St. Louis, MO",    
    subtitle = "Christopher Prener, Ph.D.",
    caption = "Data via the CDC"
  )
```

We can see from the map that elevated diabetes rates, sometimes approaching a quarter of the population, are concentrated in North St. Louis.

We'll fnish this section of the lab by saving the map to our results section:

```{r save-map}
ggsave(here("results", "stl-diabetes.png"), dpi = 300)
```

